---
title: "04-O-ana-Heatmap"
author: "Patrick Green and Ariel Marcy"
date: "01-04-2019"
output: html_document
---

### Plot PC1 Min/Max
To make the landmark variation heatmaps, we use Dr Thomas Guillerme's package `landvR` to see how specific landmarks vary along PC1. 

```{r}
library(png)  # needed to work with screenshots
library(geomorph)
library(dplyr)
library(stringr)
library(stringi)
library(devtools)
library(dispRity)
library(landvR)

# Load libraries (and install if necessary) for landvR package
if(!require(devtools)) install.packages("devtools")
if(!require(dispRity)) install.packages("dispRity")
if(!require(landvR)) install_github("TGuillerme/landvR")
```

## Prepare shape data for analyses in _geomorph_
Viewbox exports the data as a 2D array with specimens as columns and X, Y, and Z coordinates as rows for landmark. For _geomorph_, we need to convert this into a 3D array where each specimen has its own 2D array of 3 columns for XYZ coordinates and a row for each landmark. In other words, a 3D array has dimensions (p x k x n), where p is the number of landmarks, k is the dimension (2 or 3), and n is the number of specimens.

We also need to extract the metadata contained in the specimen filenames (exported as column names) and preserve their order so that the metadata matches the order of specimens in the shape data.

### Convert to 3D array
Here, we convert the coordinates into a 3D array for _geomorph_. Importantly, we need to remove the 1st column of landmark names. Following this is a 3D check of the landmarked positions. 
```{r}
data.raw <- read.csv("../1_Data (original data)/1_Raw/02_O_ana_-_Landmark_Data.csv", header = T)

data <- data.raw[, 2:(dim(data.raw)[2])]  # removes LM names for geomorph
data <- t(data)
A_ERR <- arrayspecs(data, 135, 3,sep = NULL)  # convert to 3D array w 405 pts & 3D coords

plot3d(A_ERR[,,1], asp=FALSE) #Check the landmarks have come through ok
```

## Extract metadata table from specimen filenames in shape array
The coordinate data is just shape data unless we can extract any information the landmarker put in each specimen's filename. Starting here, we make a metadata table with specimens in the same order as their coordinates for future analyses. 

### Fix naming of replicate filenames (included to measure user error)
Interspersed throughout the data collection process are ten landmarked specimens, that have been landmarked twice. Their specimen names included "_ERR", however, not always in the same place in the filename. 

We correct the naming scheme to be machine-readable and in the process create a list of specimen names which were landmarked twice. 
```{r}
# Detect replicates
names <- dimnames(A_ERR)[[3]]
isrep <- which(str_detect(names, "_ERR"))

# Removing Replicates
specimens <- c(1:length(dimnames(A_ERR)[[3]]))
notrep <- is.na(match(specimens, isrep))
A <- A_ERR[ , ,notrep]
```

### Run function to scrape filenames for metadata
The `WriteMetadata()` function makes a metadata table in the same order of specimens in the shape array, A, created from Viewbox's coordinate data. The function reads the specimen filenames from the 3D array and uses a character vector, cols, to names the attributes separated by "_" in the specimen filenames. It returns a matrix of data ordered by specimen with columns for each attribute.

## Remove "burn-in" specimens from dataset to reduce user error
First, to reduce user error, Patrick Green has landmarked 10 specimens to familiarize himself with the template and re-landmarked those 10 later on. This is close to the standard "burn-in" number recommended by GMM experts like _geomorph_ co-author, Dr Emma Sherratt (see her guides [here](http://www.emmasherratt.com/morphometrics/)).

In this dataset, the filenames look like: O_ana_MV_C12345_non for genus, species, museum code, catalog number, replicate status. Therefore, the corresponding cols vector is as below:
```{r}
WriteMetadata <- function(threeD.array, cols) {
        # Makes metadata table from specimen filenames for shape coordinates.
        #
        # Args:
        #    threeD.array: 3D array (p x k x n), where p is the number of landmarks, k is the dimension (3), and n is the number of specimens. Assumes the 1st column of landmark names has been removed. 
        #    cols: a character vector of column names of length n-1, where n is the number of underscores separating metadata information in the specimen filenames. Assumes filenames contain information in the same order and the appropriate names are given in this order in cols. 
        #
        # Returns: 
        #    A dataframe containing the metadata for each specimen in the same order as specimens in the 3D array of shape data.
        
        # Remove 'ind' that bilat.symmetry() appends to specimen names
        names <- gsub("ind", "", dimnames(threeD.array)[[3]])
        names <- gsub("_non","",names)
        names <- gsub("_ERR", "",names)
        
        # Convert name vectors into data frame
        categories <- strsplit(names, "_") 
        my.classifiers <- matrix(unlist(categories), ncol = length(cols), byrow = T) 
        colnames(my.classifiers) <- cols
        sp.info <- as.data.frame(my.classifiers)
        
        return(sp.info)
}

cols = c("Genus", "Species","Museum", "CatNum")
sp.info <- WriteMetadata(A, cols)

# Add special columns (requires 'Genus' 'Species' & 'CatNum')
sp.info$Taxa <- paste(str_sub(sp.info$Genus, 1, 1), str_sub(sp.info$Species, 1, 3), sep = "_")  # turns Ornithorhynchus anatinus into O_ana
sp.info$All <- with(sp.info, interaction(Genus, Species, Museum, CatNum, sep = "_"))  # required to run geomorph's bilat.symmetry()
```

### Create shape matrices for each patch protocol
```{r}
# Extract landmark point names
pt.names <-as.vector(data.raw[, 1])  # gets 3D point names from Viewbox
no.XYZ <- str_sub(pt.names, 3,-1)  # removes XYZ appended by Viewbox
no.XYZ.2 <- str_remove(pt.names, " [XYZ]")
u.pt.names<- unique(no.XYZ)  # returns unique point and curve names
```

### Run GPA and bilateral symmetry
Because we already slid our curve semi-landmarks and patch semi-landmarks in Viewbox, we pass their coordinates to _geomorph_ as if they are fixed landmarks. This avoids "over-sliding" the shape data. 
```{r}
Y <- gpagen(A, Proj = TRUE, ProcD = TRUE)
```


### Set up
The below was adapted from a [vignette written by Dr Guillerme](https://cdn.rawgit.com/TGuillerme/landvR/8a6a6bd5/inst/vignettes/Landmark_partition_test.html).
```{r}
# Find mean configuration - i.e. mean shape in dataset
consensus <- (select.procrustes(Y, selector = mean))[[1]]

# Measure spherical coordinates differences from the mean for each specimen for each landmark
diff.from.mean <- coordinates.difference(coordinates = Y$coords, reference = Y$consensus, type = "spherical")

# Ordinate the data
twoD.shape <- two.d.array(Y$coords)
ordination <- stats::prcomp(twoD.shape)

# Force the symmetric component of shape into class "gpagen" (required for variation.range)
gpagen.shape <- list()
gpagen.shape$coords <- Y$coords
gpagen.shape$consensus <- Y$consensus
class(gpagen.shape) <- "gpagen"

# Measure extremes of variation from mean on PC1
PC1.var <- variation.range(gpagen.shape, return.ID = FALSE, axis = 1, ordination = ordination, type = "spherical")

# Wrap specimens on the tangent space
wrap.PCA <- plotTangentSpace(Y$coords, verbose = FALSE)

# Select extreme specimens (the ones that make the warp-meshes)
hypothetical.1 <- wrap.PCA$pc.shapes[[1]]
hypothetical.2 <- wrap.PCA$pc.shapes[[2]]
```

## Plot the range of variation along PC1 using a heat color scheme
`3d` commands allow precise control of 3D visualization to get dorsal and lateral views

ERROR: rgl.snapshot() can't save the png files to the Data/Results folder so the screenshots end up in the Analysis folder
```{r}
# Dorsal view
open3d(windowRect = c(0, 0, 600, 600))
PC1.var.plot <- procrustes.var.plot(hypothetical.1, hypothetical.2, col = heat.colors, col.val = PC1.var[, "radius"], labels = F)
view3d(phi = 0)  # flip cranium
rgl.snapshot("PC1heatmap_dor.png")  # save screenshot for multipanel plot

# Lateral view
open3d(windowRect = c(0, 0, 600, 600))
PC1.var.plot <- procrustes.var.plot(hypothetical.1, hypothetical.2, col = heat.colors, col.val = PC1.var[, "radius"], labels = F)
view3d(phi = 90)
rgl.snapshot("PC1heatmap_lat.png")
```

### Prep images for multi-panel figure
This step assumes you have ImageMagik installed on your computer
```{r}
system("mogrify PC1heatmap_*.png -trim *.png")  # removes whitespace (requires ImageMagik)
dor <- readPNG("PC1heatmap_dor.png")
lat <- readPNG("PC1heatmap_lat.png")
```

## Make multipanel Figure 3
ERROR: memory issues when trying to add legend within setEPS. Works when lines 86-103 are run through the console, plot is "zoomed" and then screenshot from that window
```{r}
setEPS()  # sets up plot export
postscript("../Data/Results/Figure3_Heatmaps.eps")
layout(matrix(c(1,1,2,2), 2, 2, byrow = TRUE))  # 2 rows, 2 columns
par(mar=c(0, 0, 0, 0))

# Plot dorsal view
plot(c(0, dim(dor)[2]), c(0, dim(dor)[1]), type = "n", axes = F, xlab = "", ylab = "", asp = T) 
rasterImage(dor, 0, 0, dim(dor)[2], dim(dor)[1])
text(0, dim(dor)[1]-10, "a", cex = 2)

# Plot lateral view with same dimensions as above
plot(c(0, dim(dor)[2]), c(0, dim(dor)[1]), type = "n", axes = F, xlab = "", ylab = "", asp = T)
rasterImage(lat, 0, 0, dim(dor)[2], dim(dor)[1])
text(0, dim(dor)[1]-10, "b", cex = 2)

# Make gradient legend
legend_image <- as.raster(matrix(heat.colors(20), ncol = 1))
rasterImage(legend_image, 440, 220, 465, 160)
text(393, 213, "High variation", col = "dark grey", cex = 1.4)
text(420, 168, "Low", col = "dark grey", cex = 1.4)
dev.off()
```